---
  html_document:
    toc: true
    highlight: zenburn
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='')
```

# Overview

```{r qc-setup}
library(ggplot2)
library(reshape)
library(gplots)
library(edgeR)
library(CHBUtils)
library(pheatmap)
project_summary = "/Users/rory/cache/agnieszka-singlecell/results/project-summary.csv"
counts_file = "/Users/rory/cache/agnieszka-singlecell/results/combined.counts"
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442",
"#0072B2", "#D55E00", "#CC79A7")
summarydata = data.frame(read.table(project_summary, header=TRUE, sep=","), row.names="Name", check.rows=FALSE)
summarydata$Name = rownames(summarydata)
summarydata = summarydata[order(summarydata$Name),]
counts = read.table(counts_file, header=TRUE, row.names="id", check.names=FALSE)
counts = counts[, order(colnames(counts))]
colnames(counts) = gsub(".counts", "", colnames(counts))
# this is a list of all non user-supplied metadata columns that could appear
known_columns = c("Name", "X.GC", "Exonic.Rate", "Sequences.flagged.as.poor.quality",
    "rRNA.rate", "Fragment.Length.Mean", "Intronic.Rate", "Intergenic.Rate",
    "Mapping.Rate", "Quality.format", "Duplication.Rate.of.Mapped", "Mapped",
    "rRNA", "Sequence.length", "Transcripts.Detected", "Mean.Per.Base.Cov.",
    "Genes.Detected", "Unique.Starts.Per.Read", "unique_starts_per_read",
    "complexity", "X5.3.bias")
```

```{r heatmap-function}
get_heatmap_fn = function(summarydata) {
    # return the pheatmap function with or without metadata
    metadata = summarydata[, !colnames(summarydata) %in% known_columns, drop=FALSE]
    if(ncol(metadata) == 0) {
       return(pheatmap)
    }
    else {
    # rownames(metadata) = summarydata$Name
    heatmap_fn = function(data, ...) {
        pheatmap(data, annotation=metadata, ...)
    }
    return(heatmap_fn)
}}
heatmap_fn = get_heatmap_fn(summarydata)
```

# Quality control metrics

## Mapped reads
```{r mapped-plot}
ggplot(summarydata, aes(x=Name, y=Mapped)) +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    geom_bar(stat="identity") +
    ylab("mapped reads") + xlab("")
```

## Genomic mapping rate
```{r mapping-rate-plot}
ggplot(summarydata, aes(x=Name, y=Mapping.Rate)) +
    geom_bar(stat="identity") +
    ylab("mapping rate") + xlab("") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90))
```

## Number of genes detected
```{r genes-detected-plot}
dd = data.frame(Name=names(counts), Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Name, y=Genes.Detected)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("")
```

## Gene detection saturation
```{r saturation-plot}
dd = data.frame(Mapped=summarydata$Mapped, Genes.Detected = colSums(counts > 0))
ggplot(dd, aes(x=Mapped, y=Genes.Detected)) +
    geom_point() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("genes detected") + xlab("reads mapped")
```

## Exonic mapping rate
```{r exonic-mapping-plot}
ggplot(summarydata, aes(x=Name, y=Exonic.Rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("exonic mapping rate") + xlab("")
```

## rRNA mapping rate
```{r rRNA-rate-plot}
ggplot(summarydata, aes(x=Name, y=rRNA.rate)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("rRNA rate") + xlab("")
```

## Estimated fragment length of paired-end reads
```{r fragment-length-plot}
ggplot(summarydata, aes(x=Name, y=Fragment.Length.Mean)) +
    geom_bar(stat="identity") +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) +
    ylab("fragment length") + xlab("")
```

## Boxplot of log10 counts per gene
```{r boxplot-raw}
melted = melt(counts)
colnames(melted) = c("sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Boxplot of log10 TMM-normalized counts per gene
Trimmed mean of M-values (TMM) normalization is described
[here](http://genomebiology.com/2010/11/3/R25)

Robinson, M. D., & Oshlack, A. (2010). A scaling normalization method for differential expression analysis of RNA-seq data. Genome Biology, 11(3). doi:10.1186/gb-2010-11-3-r25

```{r boxplot-normalized}
y = DGEList(counts=counts)
y = calcNormFactors(y)
normalized_counts = cpm(y, normalized.lib.sizes=TRUE)
melted = melt(normalized_counts)
colnames(melted) = c("gene", "sample", "count")
melted$sample = factor(melted$sample)
melted = melted[order(melted$sample),]
melted$count = log(melted$count)
ggplot(melted, aes(x=sample, y=count)) + geom_boxplot() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Density of log10 TMM-normalized counts
```{r density-normalized}
ggplot(melted, aes(x=count, group=sample)) +
    geom_density() +
    theme_bw(base_size=10) +
    theme(panel.grid.major = element_line(size = .5, color = "grey"),
          axis.text.x = element_text(angle=90)) + xlab("")
```

## Correlation (Pearson) heatmap of TMM-normalized counts
```{r pearson-heatmap-normalized}
heatmap_fn(cor(normalized_counts, method="pearson"))
```

## Correlation (Spearman) heatmap of TMM-normalized counts
```{r spearman-heatmap-normalized}
heatmap_fn(cor(normalized_counts, method="spearman"))
```

## MDS plot of TMM-normalized counts
```{r mds-normalized}
mds(normalized_counts, k=length(colnames(normalized_counts)) - 1)
```

## Heatmap of top 30 most expressed genes
```{r top-count-genes, results='asis'}
select = order(rowMeans(counts),decreasing=TRUE)[1:30]
heatmap_fn(counts[select,])
```

## Heatmap by concordance correlation coefficient
http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004075

```{r propcor-heatmap}
propcor = function(x, y) {
    x = log(x + 0.1)
    y = log(y + 0.1)
    num = 2 * cov(x, y)
    denom = var(x) + var(y)
return(num/denom)}

do_propcor = function(x) {
     mat = list()
     for(i in seq_len(ncol(x))) {
         for(j in seq_len(ncol(x))) {
        x2 = x[, i]
        y2 = x[, j]
        mat = c(mat, propcor(x2, y2)) } }
    mat = unlist(mat)
    mat = matrix(mat, ncol(x), ncol(x))
    colnames(mat) = colnames(x)
    rownames(mat) = colnames(x)
    return(mat)}

heatmap_fn(do_propcor(normalized_counts))
```


## Filter out unexpressed genes and samples with low total counts
```{r filter-low-counts}
condition = "cell"
keep_cells = colSums(counts) > 1e4
counts = counts[rowSums(counts) > 0,]
counts = counts[, keep_cells]
summarydata = summarydata[colnames(counts),]
groups = summarydata[,condition]
```

```{r scde, eval=!file.exists("scde.table")}
library(scde)
n.cores = 1
o.ifm = scde.error.models(counts=counts, groups=groups, n.cores=n.cores,
                          threshold.segmentation=T, save.crossfit.plots=F,
                          save.model.plots=F, verbose=1)
valid.cells = o.ifm$corr.a > 0
o.ifm = o.ifm[valid.cells,]
groups = groups[valid.cells]
counts = counts[valid.cells,]
o.prior = scde.expression.prior(models=o.ifm, counts=counts, length.out=400,
                                show.plot=F)
ediff <- scde.expression.difference(o.ifm, counts, o.prior, groups=groups,
                                    n.randomizations=100,
                                    n.cores=n.cores, verbose=1)
ediff$pvalue = pnorm(-(abs(ediff$Z))) * 2
ediff$padj = p.adjust(ediff$pvalue)
ediff = ediff[order(ediff$pvalue),]
save(o.ifm, groups, counts, o.prior, ediff, file="scde.RData")
```

```{r add-symbol}
load("scde.RData")
library(biomaRt)
mouse = useMart("ensembl", dataset="mmusculus_gene_ensembl")
conversions = getBM(attributes=c("ensembl_gene_id", "mgi_symbol"), mart=mouse)
matches = match(rownames(ediff), conversions$ensembl_gene_id)
ediff$symbol = conversions$mgi_symbol[matches]
write.table(ediff, file="singlecell_de.tsv", sep="\t", quote=FALSE,
            row.names=TRUE, col.names=TRUE)
```

```{r seurat}
library(Seurat)
nbd.data = data.frame(log(counts + 1))
cnames = paste(colnames(nbd.data), groups, sep="_")
colnames(nbd.data) = cnames
nbt = new("seurat", raw.data=nbd.data)
nbt= setup(nbt, project="NBT", min.cells = 3, min.genes = 1000, is.expr=1,
           names.delim="_", names.field=2)
```

Below we identify highly variable genes and use those to do PCA on the samples.

```{r identify-variable-genes}
nbt = mean.var.plot(nbt, y.cutoff=2, x.low.cutoff=2, fxn.x=expMean,
                    fxn.y=logVarDivMean)
length(nbt@var.genes)
```

## Below we calculate the weights for the first 10 components to pick out
## the ones that explain most of the variance, then look at the separation along
## those components.

## ```{r pca-jackstraw}
## nbt = pca(nbt, do.print=FALSE)
## nbt = jackStraw(nbt, num.replicate=200, do.print=FALSE)
## jackStrawPlot(nbt, PCs=1:6)
## pca.plot(nbt, 1, 2, pt.size=2)
## viz.pca(nbt, 1:2)
## pca.plot(nbt, 3, 4, pt.size=2)
## pca.plot(nbt, 5, 6, pt.size=2)

## nbt=run_tsne(nbt,dims.use = 1:4,max_iter=2000)
## tsne.plot(nbt,pt.size = 3)
## ```

## ```{r scde-heterogeneity, eval=!file.exists("knn.RData")}
## knn <- knn.error.models(counts, k = ncol(counts)/4, n.cores = 2, min.count.threshold = 2, min.nonfailed = 5, max.model.plots = 10)
## save(knn, file="knn.RData")
## ```

## ```{r varinfo}
## load("knn.RData")
## varinfo <- pagoda.varnorm(knn, counts = counts, trim = 3/ncol(counts), max.adj.var = 5, n.cores = 1, plot = TRUE)
